preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-08 Manuscript/manuscript.Rmd
Fixes for all ??? that MD could fix

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome [@dixon2012topological;@rao20143d;@sexton2012three;@chen2016novo;@filippova2014identification;@shin2016topdom;@weinreb2016identification], less so for chromatin loops [@salameh2019supervised;@kai2018predicting]. [??? Check refs, the current ones are suboptimal. Check SpectralTAD paper for refs about TAD callers, use https://github.com/mdozmorov/HiC_tools#loop-callers for loop callers] However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin]. Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised machine learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at much higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability]

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? Is there other resolution? ??? TAD and loop boundaries should be distinguished, to describe when we have all results] [??? Define that ARROWHEAD data constitute the main results]

@@

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? DEfine all indices, what is N, p+1, i?]

@@

![**???[needs updating with new framework, takes longest, do last] Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for GM12879 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).](Figures/figure3_performances.png)

@@

We sought to further optimize our TAD boundary region prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin].

@@

Using our optimally built random forest model, we attempted to overcome resolution limitations of Hi-C data by predicting TAD [??? Check throughout the manuscript - we may need to define "interacting domains" referring to TADs and loops] boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified [??? We need pre-trained models and understanding how they perform on different cell lines] model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location ad base resolution (see Methods).

@@

Our random forest framework yielded several interesting observations. First, we showed that RF models built using *distance*-type predictors out performed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational]. [??? Add more REFs] We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique. The unaddressed class imbalance problem in previous studies [??? Add REFs] We showed that information about only four transcription factors (SMC3, Rad21, CTCF, Znf143) is necessary and sufficient transcription for accurate boundary prediction, out-performing histone modification- and BroadHMM-built models. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. Although our observations contradict a previous study showing histone marks alone were effective in predicting TAD boundaries [@huang2015predicting], our findings are backed by a large body of literature [??? Add REFs]. This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting boundary regions.

>>>>

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), less so for chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as TAD callers for brevity, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among TAD callers is that they disregard prior knowledge about functional genomic elements associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised machine learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at much higher resolutions.

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability. For MD, last part when ready to write an abstract]

@@

Our framework is resolution-specific and uses called TAD or loop boundaries as ground truth to establish TAD(loop)-boundary and non-TAD(loop) boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines obtained from GEO GSE63525 (Supplementary Table 1) [??? Update table and methods with relevant data only, not individual samples]. TADs were subsequently called using the ARROWHEAD pipeline from the juicer toolbox [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. For loop boundaries, we opted to use boundaries obtained by the Peakachu prediction tool for the GM12878 and K562 cell lines at 10 kb resolution [@salameh2019supervised]. [??? TAD and loop boundaries should be distinguished, to describe when we have all results. For MD] [??? Define that ARROWHEAD data constitute the main results. For MD]

@@

![**A diagram of the machine learning framework.** Step 1 defines the chromosome-specific predictor-response matrix **A** of $N$ genomic bins (rows), $p$ genomic annotations, and one response vector $Y$ ($p+1$ columns). Resolution-specific data (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)) for the $n-1$ autosomal chromosomes. Step 2 is to reserve the predictor-response matrix for the holdout chromosome $i$ as the test data. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/figure2_ensemble_framework.png) [??? Define all indices, what is N, p+1, i?]

@@

![**???[needs updating with new framework, takes longest, do last - ??? Is this current?] Distance-type predictors combined with random under-sampling yielded optimal TAD boundary region predictive performance for GM12879 ARROWHEAD TAD boundaries.** Balanced accuracy is compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple).](Figures/figure3_performances.png)

@@

We sought to further optimize our TAD boundary region prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].

@@

Using our optimally built random forest model trained on ARROWHEAD/Peakachu boundaries, we attempted to predict TAD/loop boundaries at base resolution. Intuitively, the predictor-response space is built on a base-resolution level, and the model trained on a bin-level space predicts the probability of each base being a boundary. Our method, referred to as *preciseTAD*, leverages a user-specified [??? We need pre-trained models and understanding how they perform on different cell lines] model with density-based hierarchical clustering (DBSCAN) and a scalable partitioning around mediods clustering (CLARA) to predict the boundary location ad base resolution (see Methods).

@@

Our random forest framework yielded several interesting observations. First, we showed that RF models built using *distance*-type predictors outperformed models built on previously published feature engineering techniques including signal strength, overlap counts and overlap percents [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa; @kai2018predicting]. We further demonstrated that class imbalance hinders boundary prediction, but can be effectively addressed by a simple random under-sampling technique. The unaddressed class imbalance problem in previous studies [@mourad2016computational; @Moore:2015aa; @Al-Bkhetan:2018aa] We showed that information about only four transcription factors (SMC3, Rad21, CTCF, Znf143) is necessary and sufficient transcription for accurate boundary prediction, outperforming histone modification- and BroadHMM-built models [@Moore:2015aa; @Al-Bkhetan:2018aa]. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Interestingly, the same transcription factors accurately predicted both TAD and loop boundaries, suggesting similarity of the mechanisms of TAD and loop formation. Although our observations contradict a previous study showing histone marks alone were effective in predicting TAD boundaries [@huang2015predicting], our findings are backed by a large body of literature [??? Add REFs]. This suggested that the random forest model, when tuned and feature engineered correctly, is highly effective in predicting boundary regions.
